---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/README-"
)
```

# ALA4R

[![Travis-CI Build Status](https://travis-ci.org/AtlasOfLivingAustralia/ALA4R.svg?branch=master)](https://travis-ci.org/AtlasOfLivingAustralia/ALA4R)
[![AppVeyor Build status](https://ci.appveyor.com/api/projects/status/mdsmv5n90gv54b1o/branch/master?svg=true)](https://ci.appveyor.com/project/raymondben/ala4r/branch/master)
[![Coverage Status](https://img.shields.io/codecov/c/github/AtlasOfLivingAustralia/ALA4R/master.svg)](https://codecov.io/github/AtlasOfLivingAustralia/ALA4R?branch=master)
[![CRAN Status](http://www.r-pkg.org/badges/version/ALA4R)](http://www.r-pkg.org/pkg/ALA4R)


The Atlas of Living Australia (ALA) provides tools to enable users of biodiversity information to find, access, combine and visualise data on Australian plants and animals; these have been made available from http://www.ala.org.au/. Here we provide a subset of the tools to be directly used within R.

ALA4R enables the R community to directly access data and resources hosted by the ALA.

The use-examples presented at the [2014 ALA Science Symposium](http://www.ala.org.au/blogs-news/2014-atlas-of-living-australia-science-symposium/) are available in the package vignette, via (in R): `vignette("ALA4R")`, or [browse it online](https://atlasoflivingaustralia.github.io/ALA4R/articles/ALA4R.html).

## Installing

Stable version from CRAN:

```{r eval=FALSE}
install.packages("ALA4R")
```

Or the development version from GitHub:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("AtlasOfLivingAustralia/ALA4R")
```

If you wish to use the `data.table` package for potentially faster loading of data matrices (optional), also do:
```{r eval=FALSE}
install.packages("data.table")
```

On Linux you will first need to ensure that `libcurl` and `v8` (version <= 3.15) are installed on your system --- e.g. on Ubuntu/Debian, open a terminal and do:

```{sh eval=FALSE}
sudo apt-get install libcurl4-openssl-dev libv8-3.14-dev
```
or install via the Software Centre.


## Usage

See the online documentation at https://atlasoflivingaustralia.github.io/ALA4R/ including the [package vignette](https://atlasoflivingaustralia.github.io/ALA4R/articles/ALA4R.html).
