---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/README-"
)
```

# ALA4R

[![Travis-CI Build Status](https://travis-ci.com/AtlasOfLivingAustralia/ALA4R.svg?branch=master)](https://travis-ci.com/AtlasOfLivingAustralia/ALA4R)
[![AppVeyor Build status](https://ci.appveyor.com/api/projects/status/g9pudc4l7053w4vn/branch/master?svg=true)](https://ci.appveyor.com/project/PeggyNewman/ala4r/branch/master)
[![codecov](https://codecov.io/gh/AtlasOfLivingAustralia/ALA4R/branch/master/graph/badge.svg)](https://codecov.io/gh/AtlasOfLivingAustralia/ALA4R)
[![CRAN Status](https://www.r-pkg.org/badges/version/ALA4R)](https://www.r-pkg.org/pkg/ALA4R)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/ALA4R)](https://cran.r-project.org/package=ALA4R)

ALA4R enables the R community to directly access data hosted by the Atlas of Living Australia (ALA).

The ALA aggregates Australian biodiversity data from a range of sources, and makes the data freely available for scientists, policy makers, industry and the general public.

**NOTE:** The next release of ALA4R will be version 2.0.0. This will be a major update to simplify the process of downloading data, make behaviour across functions consistent, and remove redundant functionality.  
In the coming weeks, you will be able to test out the new functionality in the [dev branch](https://github.com/AtlasOfLivingAustralia/ALA4R/tree/dev).
If you have any questions or suggestions, please [contact us](mailto:support@ala.org.au).

## Citing ALA4R 
To generate a citation for ALA4R run: 
```r
citation(package = "ALA4R")
```

## Installing

Stable version from CRAN:

```{r eval=FALSE}
install.packages("ALA4R")
```

Or the development version from GitHub:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("AtlasOfLivingAustralia/ALA4R")
```

On Linux you will first need to ensure that `libcurl` and `v8` (version <= 3.15) are installed on your system --- e.g. on Ubuntu/Debian, open a terminal and do:

```{sh eval=FALSE}
sudo apt-get install libcurl4-openssl-dev libv8-3.14-dev
```


## Usage

See the online documentation at https://atlasoflivingaustralia.github.io/ALA4R/ including the [package vignette](https://atlasoflivingaustralia.github.io/ALA4R/articles/ALA4R.html).
