<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{ALA4R}
-->

# ALA4R

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center")
```


This project enables the R community to directly access data and resources hosted by the Atlas of Living Australia (ALA: http://ala.org.au/). Our goal is to enable outputs (e.g. observations of species) to be queried and output in a range of standard formats.

## Installing

### Windows

First, install some packages from CRAN that are required by our package (CRAN is a network of ftp and web servers around the world that store identical, up-to-date, versions of code and documentation for R):
```{r eval=FALSE}
install.packages(c("httr","stringr","plyr","digest","RCurl","jsonlite","assertthat","sp","knitr"))
```

If you wish to use the `data.table` package for potentially faster loading of data matrices (optional), also do:
```{r eval=FALSE}
install.packages(c("data.table"))
```

Then install the ALA4R package itself from RForge:
```{r eval=FALSE}
install.packages("ALA4R",repos="http://rforge.net/",type="source")
```

### Linux

First, ensure that `libcurl` is installed on your system --- e.g. on Ubuntu, open a terminal and do:
```
sudo apt-get install libcurl4-openssl-dev
```

or install `libcurl4-openssl-dev` via the Software Centre.

Then, fire up R and install the dependencies that we need:
```{r eval=FALSE}
install.packages(c("httr","stringr","plyr","digest","RCurl","jsonlite","assertthat","sp","knitr"))
```

If you wish to use the `data.table` package for potentially faster loading of data matrices (optional), also do:
```{r eval=FALSE}
install.packages(c("data.table"))
```

Then the ALA4R package itself:
```{r eval=FALSE}
install.packages("ALA4R",repos="http://rforge.net/")
```



## Using

The ALA4R package must be loaded for each new R session:

```{r}
library(ALA4R)
```

## Customizing

Various aspects of the ALA4R package can be customized.

### Caching
ALA4R can cache most results to local files. This means that if the same code is run multiple times, the second and subsequent iterations will be faster. This will also reduce load on the ALA servers.

By default, this caching is session-based, meaning that the local files are stored in a temporary directory that is automatically deleted when the R session is ended. This behaviour can be altered so that caching is permanent, by setting the caching directory to a non-temporary location. For example, to cache your results in the Windows directory "c:\mydata\ala_cache", use:

```{r eval=FALSE}
ala_config(cache_directory=file.path("c:","mydata","ala_cache")) ## Windows
```

or for Linux:

```{r eval=FALSE}
ala_config(cache_directory=file.path("~","mydata","ala_cache")) ## Linux
```

Note that this directory must exist (you need to create it yourself).


All results will be stored in that cache directory and will be used from one session to the next. They won't be re-downloaded from the server unless the user specifically deletes those files or changes the caching setting to "refresh".

If you change the cache_directory to a permanent location, you may wish to add something like this to your .Rprofile file, so that it happens automatically each time the ALA4R package is loaded:

```{r eval=FALSE}
setHook(packageEvent("ALA4R", "attach"), function(...) ala_config(cache_directory=file.path("~","mydata","ala_cache")))
```

Caching can also be turned off entirely by:

```{r eval=FALSE}
ala_config(caching="off")
```

or set to "refresh", meaning that the cached results will re-downloaded from the ALA servers and the cache updated. (This will happen for as long as caching is set to "refresh" --- so you may wish to switch back to normal "on" caching behaviour once you have updated your cache with the data you are working on).

### User-agent string
Each request to the ALA servers is accompanied by a "user-agent" string that identifies the software making the request. This is a standard behaviour used by web browsers as well. The user-agent identifies the user requests to the ALA, helping the ALA to adapt and enhance the services that it provides. By default, the ALA4R user-agent string is set to "ALA4R" plus the ALA4R version number, R version, and operating system (e.g. "ALA4R 0.16 (R version 3.0.2 (2013-09-25)/x86_64-pc-linux-gnu)"). 

*NO* personal identification information is sent. You can see all configuration settings, including the the user-agent string that is being used, with the command:

```{r eval=FALSE}
ala_config()
```

### Debugging
If things aren't working as expected, more detail (particularly about web requests and caching behaviour) can be obtained by setting the `verbose` configuration option:

```{r eval=FALSE}
ala_config(verbose=TRUE)
```


